Parallel algorithm for solving multicriterial optimization problems using elements of machine learning

### 2025-06-18 2025/RSCD_MCO/RSD-2025.tex
sort ref.

The problems of choosing globally optimal solutions arise when conducting a wide class of researches. To find the best options, an efficiency criterion is determined, the values   of which depend on the parameters from the permissible domain. The criterion itself can often be presented not in an analytical form, but only as a procedure for calculating its values, i.e. in the form of a ``black box'' and additionally it can be multiextremal \cite{Miettinen1999,Ehrgott2005,Pardalos2017,Strongin2000,Sergeyev2013}. For a more accurate description of the problem being solved, several (generally speaking, contradictory) optimized criteria can be taken.

Multiextremal problems of global optimization are computationally complicated because the globally optimal solution, because in the general case, to claim that the solution is globally optimal, you need to compare it with the values of the objective function at all other points in the search area. When solving such problems using the scanning methods based on random or uniform grids, the number of calculations increases exponentially with the growth of dimensionality. It is possible to solve the considered problems using more efficient algorithms that can be divided into two large classes: deterministic \cite{Miettinen1999,Ehrgott2005,Pardalos2017,Konnov2025,Evtushenko2014} and non-deterministic (meta-heuristic) methods \cite{DPA02,Durillo2010,Mostaghim2007,Nebro2009,RC05,Zitzler2001}.

@@

Deterministic algorithms guarantee an estimate of the globally optimal solution with a given accuracy, but in their basic implementations they can perform redundant calculations. To reduce the number of trials in this class of algorithms, heuristics are often used that reduce the amount of computations \cite{Konnov2025,AGP_ML}. Among effective deterministic algorithms for solving global optimization problems, the information-statistical family of global search methods \cite{Strongin2000,Sergeyev2013} can be considered. Initially, these algorithms were proposed for solving scalar optimization problems, subsequently, the algorithms were successfully developed to solve multicriterial problems of global search \cite{Konnov2025,MCO_ML_2023}.

@@

As a particular solution to the MCO problem, any effective option can be considered in which it is impossible to reduce the values of all criteria at once $f_i(y)$, $1 \leq i \leq s$, by selection of parameter values $y \in D$. In general case, when solving the MCO problem, it may be necessary to find the entire set of Pareto-optimal options \cite{Miettinen1999,Ehrgott2005,Pardalos2017}.

@@

\textbf{Scalarization of efficiency criteria.} It is possible to find the Pareto set either solving the problem in the original formulation \cite{Evtushenko2014,DPA02,Durillo2010,Mostaghim2007,Nebro2009,RC05,Zitzler2001}, or reducing the original problem to a series of scalar problems to find effective solutions \cite{Pardalos2017,Konnov2025,Gergel2019_2,Gergel2018,GergelKozinov2020}.

In the framework of the developed approach \cite{Konnov2025}, the second option with scalarization of criteria is used. To build a good estimate of the Pareto set, a family of problems
\begin{equation}
\label{eq:04}
\min {\varphi_j (y)}=F(\lambda_j,y),F(\lambda_j,y) \in \{F(\lambda_1,y),F(\lambda_2,y),\dots,F(\lambda_\tau,y)\},y\in D,
\end{equation}
is solved,  where $F(\lambda_j,y)$, $1 \leq j \leq \tau$, is a multiextremal function, obtained on the base of scalarization of criteria $f_i(y)$, $1 \leq i \leq s$. Here $\lambda_j=\{\lambda_{i,j}\}$, $1 \leq i \leq s$, $1 \leq j \leq \tau$, are significance indicators for each criterion, s is the number of criteria, $\tau$ is the number of scalar optimization problems. There are different ways to build functions $F(\lambda_j,y)$ \cite{Miettinen1999,Konnov2025,Marler2004}.  
In the proposed algorithm we used fixed value $\tau = 50$. The values $\lambda_j$, $1 \leq j \leq \tau$, are selected uniformly within the set of values satisfying the condition (\ref{eq:051})
\begin{equation}
\label{eq:051}
\sum_{i=1}^s {\lambda_{i,j}} = 1, 0 \leq \lambda_{i,j} \leq 1.
\end{equation}

@@

\textbf{Dimensionality reduction.} For solving optimization problems (\ref{eq:04}) the dimensionality reduction on the base of Peano curves $y(x)$ that map unambiguously and continuously the segment $[0,1]$ to the $N$-dimensional hypercube $D$ is used \cite{Konnov2025,Gergel2019_2,GergelKozinov2020}. As a result of this reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional subproblems
\begin{equation}
\label{eq:06}
\min_x {\varphi_j(y(x))} = \min_y {\varphi_j(y)}, x \in [0,1], y \in D, 1 \leq j \leq \tau
\end{equation}

@@

\textbf{Global search algorithm.} Information-statistical algorithm of global search (shortly GSA) \cite{Konnov2025,Gergel2019_2,Gergel2018,GergelKozinov2020,Strongin2000,Sergeyev2013} is used to solve problems (\ref{eq:06}). The computational scheme of GSA is as follows. At each iteration of the global search, one trial (calculation of the value of the optimized function $\varphi_j (x)$ from (\ref{eq:06})) is executed. In the first iteration, two tests are carried out at the ends of the interval $x^0=0$,$x^1=1$. Let $k$, $k>2$, iterations of the global search be performed. The selection of the next test point is based on the following rules.

@@

Within item 1 the following sequence of actions is performed. When task (\ref{eq:06}) is changed, new coefficients $\lambda_{j+1}$ from (\ref{eq:04}) are selected. Further, scalarization of the vector criterion (\ref{eq:04}) and dimensionality reduction (\ref{eq:06}) transform SIS from (\ref{eq:14}) to the form used in the optimization process (\ref{eq:08})-(\ref{eq:13}) for the current problem (\ref{eq:04}). Transformation uses already calculated criteria values and new coefficients $\lambda_{j+1}$. As a consequence, the recount does not take long time. Thus, information about the calculated criteria is not lost when changing the task statement, but is reused. This approach has shown its effectiveness in solving many problems \cite{Konnov2025,Gergel2018,GergelKozinov2020}.

@@

From the figure, it can be seen that the function $R_{PS}(i)$ for each of the considered variants differs in the quality of the evaluation of Pareto sets. When choosing a machine learning model, it should be taken into account that building a model with a good description of the domain may require more computing resources. An analysis of the impact of the selected model on the quality of the resulting solution is presented in Section \ref{sec4} with the results of computational experiments. Note that numerous computational experiments have shown that changing the method of calculating characteristic (\ref{eq:15}) leads to a significant increase in the efficiency of the algorithm as a whole \cite{Konnov2025,MCO_ML_2023}.

@@

As a criterion for the quality of the found Pareto set, the indicator HV was applied. Higher value of the indicator corresponds to better solution. 
HV index is a traditional quality metric when approximating a Pareto set \cite{Evtushenko2014,pymoo,AGP_ML}.
To assess the performance of the algorithm, the number of trials carried out when solving the problem are given. However, when solving complex optimization problems, one trial can take a significant time. In such conditions, the reduction of the number of trials reduces the total computation time in general. In the presented experimental results, the number of trials spent on solving each problem and the HV indicators were averaged according to the number of solved problems.

>>>>

The problems of choosing globally optimal solutions arise when conducting a wide class of researches. To find the best options, an efficiency criterion is determined, the values   of which depend on the parameters from the permissible domain. The criterion itself can often be presented not in an analytical form, but only as a procedure for calculating its values, i.e. in the form of a ``black box'' and additionally it can be multiextremal \cite{Ehrgott2005,Miettinen1999,Pardalos2017,Sergeyev2013,Strongin2000}. For a more accurate description of the problem being solved, several (generally speaking, contradictory) optimized criteria can be taken.

Multiextremal problems of global optimization are computationally complicated because the globally optimal solution, because in the general case, to claim that the solution is globally optimal, you need to compare it with the values of the objective function at all other points in the search area. When solving such problems using the scanning methods based on random or uniform grids, the number of calculations increases exponentially with the growth of dimensionality. It is possible to solve the considered problems using more efficient algorithms that can be divided into two large classes: deterministic \cite{Ehrgott2005,Evtushenko2014,Konnov2025,Miettinen1999,Pardalos2017} and non-deterministic (meta-heuristic) methods \cite{DPA02,Durillo2010,Mostaghim2007,Nebro2009,RC05,Zitzler2001}.

@@

Deterministic algorithms guarantee an estimate of the globally optimal solution with a given accuracy, but in their basic implementations they can perform redundant calculations. To reduce the number of trials in this class of algorithms, heuristics are often used that reduce the amount of computations \cite{AGP_ML,Konnov2025}. Among effective deterministic algorithms for solving global optimization problems, the information-statistical family of global search methods \cite{Sergeyev2013,Strongin2000} can be considered. Initially, these algorithms were proposed for solving scalar optimization problems, subsequently, the algorithms were successfully developed to solve multicriterial problems of global search \cite{MCO_ML_2023,Konnov2025}.

@@

As a particular solution to the MCO problem, any effective option can be considered in which it is impossible to reduce the values of all criteria at once $f_i(y)$, $1 \leq i \leq s$, by selection of parameter values $y \in D$. In general case, when solving the MCO problem, it may be necessary to find the entire set of Pareto-optimal options \cite{Ehrgott2005,Miettinen1999,Pardalos2017}.

@@

\textbf{Scalarization of efficiency criteria.} It is possible to find the Pareto set either solving the problem in the original formulation \cite{DPA02,Durillo2010,Evtushenko2014,Mostaghim2007,Nebro2009,RC05,Zitzler2001}, or reducing the original problem to a series of scalar problems to find effective solutions \cite{Gergel2019_2,Gergel2018,GergelKozinov2020,Konnov2025,Pardalos2017}.

In the framework of the developed approach \cite{Konnov2025}, the second option with scalarization of criteria is used. To build a good estimate of the Pareto set, a family of problems
\begin{equation}
\label{eq:04}
\min {\varphi_j (y)}=F(\lambda_j,y),F(\lambda_j,y) \in \{F(\lambda_1,y),F(\lambda_2,y),\dots,F(\lambda_\tau,y)\},y\in D,
\end{equation}
is solved,  where $F(\lambda_j,y)$, $1 \leq j \leq \tau$, is a multiextremal function, obtained on the base of scalarization of criteria $f_i(y)$, $1 \leq i \leq s$. Here $\lambda_j=\{\lambda_{i,j}\}$, $1 \leq i \leq s$, $1 \leq j \leq \tau$, are significance indicators for each criterion, s is the number of criteria, $\tau$ is the number of scalar optimization problems. There are different ways to build functions $F(\lambda_j,y)$ \cite{Konnov2025,Marler2004,Miettinen1999}.  
In the proposed algorithm we used fixed value $\tau = 50$. The values $\lambda_j$, $1 \leq j \leq \tau$, are selected uniformly within the set of values satisfying the condition (\ref{eq:051})
\begin{equation}
\label{eq:051}
\sum_{i=1}^s {\lambda_{i,j}} = 1, 0 \leq \lambda_{i,j} \leq 1.
\end{equation}

@@

\textbf{Dimensionality reduction.} For solving optimization problems (\ref{eq:04}) the dimensionality reduction on the base of Peano curves $y(x)$ that map unambiguously and continuously the segment $[0,1]$ to the $N$-dimensional hypercube $D$ is used \cite{Gergel2019_2,GergelKozinov2020,Konnov2025}. As a result of this reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional subproblems
\begin{equation}
\label{eq:06}
\min_x {\varphi_j(y(x))} = \min_y {\varphi_j(y)}, x \in [0,1], y \in D, 1 \leq j \leq \tau
\end{equation}

@@

\textbf{Global search algorithm.} Information-statistical algorithm of global search (shortly GSA) \cite{Gergel2019_2,Gergel2018,GergelKozinov2020,Konnov2025,Sergeyev2013,Strongin2000} is used to solve problems (\ref{eq:06}). The computational scheme of GSA is as follows. At each iteration of the global search, one trial (calculation of the value of the optimized function $\varphi_j (x)$ from (\ref{eq:06})) is executed. In the first iteration, two tests are carried out at the ends of the interval $x^0=0$,$x^1=1$. Let $k$, $k>2$, iterations of the global search be performed. The selection of the next test point is based on the following rules.

@@

Within item 1 the following sequence of actions is performed. When task (\ref{eq:06}) is changed, new coefficients $\lambda_{j+1}$ from (\ref{eq:04}) are selected. Further, scalarization of the vector criterion (\ref{eq:04}) and dimensionality reduction (\ref{eq:06}) transform SIS from (\ref{eq:14}) to the form used in the optimization process (\ref{eq:08})-(\ref{eq:13}) for the current problem (\ref{eq:04}). Transformation uses already calculated criteria values and new coefficients $\lambda_{j+1}$. As a consequence, the recount does not take long time. Thus, information about the calculated criteria is not lost when changing the task statement, but is reused. This approach has shown its effectiveness in solving many problems \cite{Gergel2018,GergelKozinov2020,Konnov2025}.

@@

From the figure, it can be seen that the function $R_{PS}(i)$ for each of the considered variants differs in the quality of the evaluation of Pareto sets. When choosing a machine learning model, it should be taken into account that building a model with a good description of the domain may require more computing resources. An analysis of the impact of the selected model on the quality of the resulting solution is presented in Section \ref{sec4} with the results of computational experiments. Note that numerous computational experiments have shown that changing the method of calculating characteristic (\ref{eq:15}) leads to a significant increase in the efficiency of the algorithm as a whole \cite{MCO_ML_2023,Konnov2025}.

@@

As a criterion for the quality of the found Pareto set, the indicator HV was applied. Higher value of the indicator corresponds to better solution. 
HV index is a traditional quality metric when approximating a Pareto set \cite{AGP_ML,pymoo,Evtushenko2014}.
To assess the performance of the algorithm, the number of trials carried out when solving the problem are given. However, when solving complex optimization problems, one trial can take a significant time. In such conditions, the reduction of the number of trials reduces the total computation time in general. In the presented experimental results, the number of trials spent on solving each problem and the HV indicators were averaged according to the number of solved problems.
